# Least squares estimation

------------------------------------------------------------------------

## Least squares: the intuitive approach

[**Core Idea:**]{style="color: #FFFF00;"} Minimize the sum of squared differences between model predictions and observations

------------------------------------------------------------------------

**Mathematical Formulation:** $$\text{SSE} = \sum_{i=1}^{n} (y_i - f(t_i, \theta))^2$$

Where:

-   $y_i$ = observed value at time $t_i$
-   $f(t_i, \theta)$ = model prediction at time $t_i$
-   $\theta$ = parameter vector

------------------------------------------------------------------------

### Why squared errors?

[**Advantages:**]{style="color: #FFFF00;"}

-   [Penalizes large errors more heavily]{style="color: #FFFF00;"}
-   [Differentiable (smooth optimization)]{style="color: #FFFF00;"}
-   [Mathematically tractable]{style="color: #FFFF00;"}
-   [Gives maximum likelihood estimates when errors are normal distributed]{style="color: #FFFF00;"}

------------------------------------------------------------------------

### [Disadvantages]{style="color: tomato;"}

-   [Sensitive to outliers]{style="color: tomato;"}
-   [Assumes constant variance]{style="color: tomato;"}
-   [No probabilistic interpretation]{style="color: tomato;"}

------------------------------------------------------------------------

## Least squares example: SIR model

```{r ls-example, echo=FALSE, fig.width=10, fig.height=6}
# Load required packages
library(deSolve)
library(ggplot2)

# Define SIR model
sir_model <- function(t, y, params) {
  with(as.list(c(y, params)), {
    dS <- -beta * S * I
    dI <- beta * S * I - gamma * I
    dR <- gamma * I
    list(c(dS, dI, dR))
  })
}

# Generate synthetic data
set.seed(42)
true_params <- c(beta = 0.3, gamma = 0.1)
init_conds <- c(S = 0.99, I = 0.01, R = 0)
times <- seq(0, 365, by = 1)

# Simulate true model
out <- ode(y = init_conds, times = times, func = sir_model, parms = true_params)
true_cases <- out[,"I"] * 1000  # Scale to cases

# Add noise
observed_cases <- rpois(length(times), true_cases + 5)

# Plot
plot_data <- data.frame(
  time = times,
  observed = observed_cases,
  true_model = true_cases
)

ggplot(plot_data, aes(x = time)) +
  geom_point(aes(y = observed), color = "red", alpha = 0.7) +
  geom_line(aes(y = true_model), color = "blue", linewidth = 1) +
  labs(x = "Time (days)", y = "Number of Cases",
       title = "SIR Model: True vs Observed Data") +
  theme_minimal()
```

------------------------------------------------------------------------

## Implementing least squares

```{r ls-implementation, echo=TRUE}
#| code-line-numbers: "1-2|3-12|14-15|16-19|1-15"
# Define objective function
sse_function <- function(params, data) {
  beta <- params[1]
  gamma <- params[2]
  
  # Simulate model
  out <- ode(
    y = init_conds,
    times = data$time,
    func = sir_model,
    parms = c(beta = beta, gamma = gamma)
  )
  
  #  Extract and rescale the predicted cases
  predicted <- out[, "I"] * 1000
  # Calculate sum of squared errors
  sse <- sum((data$observed - predicted)^2)
  
  return(sse)
}
```

-------------------------------------------------------------------------

```{r ls-fitting, echo=TRUE}
#| code-line-numbers: "1-2|4-8|6|7|4-7" 
# Prepare data
data <- data.frame(time = times, observed = observed_cases)

# Test different parameter values
test_params <- expand.grid(
  beta = seq(0.1, 0.5, length.out = 10),
  gamma = seq(0.05, 0.2, length.out = 10)
)
```

--------------------------------------------------------------------------

```{r ls-fitting-2, echo=TRUE}
#| code-line-numbers: "1-2|3|4-7|8|1-7"
# Calculate SSE for each combination
for (i in 1:nrow(test_params)) {
  test_params$sse[i] <- sse_function(
    c(
      test_params[i, 1],
      test_params[i, 2]
    ),
    data
  )
}
```

--------------------------------------------------------------------------

```{r ls-results, echo=TRUE}
#| code-line-numbers: "1-2|3|4-7|1-7"
# Find minimum
best_idx <- which.min(test_params$sse)
best_params <- test_params[best_idx, ]
cat("True parameters: Beta =", true_params[1], ", Gamma =", true_params[2])
cat("Best fit parameters: Beta =", round(best_params$beta, 3), ", Gamma =", round(best_params$gamma, 3), "\n")
```

------------------------------------------------------------------------

## Strengths of least squares

**Computational Advantages:**

-   Fast and efficient
-   Well-established algorithms
-   Easy to implement
-   Good for initial parameter estimates

------------------------------------------------------------------------

## Statistical properties

-   Unbiased estimates (under certain conditions)
-   Minimum variance among linear unbiased estimators
-   Maximum likelihood when errors are normal

------------------------------------------------------------------------

## Practical benefits

-   Intuitive interpretation
-   Widely understood
-   Good starting point for more complex methods

------------------------------------------------------------------------

## Limitations of least squares

**Statistical Limitations:**

-   Limited uncertainty quantification
-   Assumes constant variance
-   Sensitive to outliers
-   No probabilistic framework

------------------------------------------------------------------------

## Practical limitations

**Practical Limitations:**

-   Parameter identifiability issues
-   No confidence intervals
-   Difficult to compare models
-   Assumes measurement error only

------------------------------------------------------------------------

## R implementation practicals {background-color="#447099" transition="fade-in"}

-   Let's turn to the [tutorials](https://github.com/jamesmbaazam/mppr_intro_to_fitting_practicals)
